CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems

### 2014-11-24 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
Updating according to Vaughn's second review. Almost all sections by Ehsan are done by this update.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4509 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the ``flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}). The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

@@

We believe the key to the good performance of the new placer cost functions is that it produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later smaller placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalize bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{bb.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

@@

\subsubsection{Effectiveness of router optimizations}
\label{sec:effectiveness_of_router_enhancements}
The router optimizations described in Section~\ref{sec:router_optimizations} extensively improve the runtime of the router lookahead algorithm. Figure~\ref{fig:lookahead_runtime} shows geomean of the router runtime for the 8 circuits for different \textit{\% wires cut} values. On average, the router runs 31 times faster and finds identical minimum channel widths for all circuits.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt.eps}
\caption{Router runtime improvement using updated lookahead function}
\label{fig:lookahead_runtime}
\end{figure}

The new router lookahead changes the priority of the routes discovered by the router. Using the new router lookahead function increases the critical path delay by 1.3\% on average as shown in Figure~\ref{fig:lookahead_fmax}.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt_fmax.eps}
\caption{The new router lookahead increases the critical path delay.}
\label{fig:lookahead_fmax}
\end{figure}

@@

Using the best CAD settings found in Section~\ref{cadSection} and the best CAD flow described in Section~\ref{sec:JaveedSection}, we analyze the impact of the key architecture parameters: \textit{\% wires cut}, \textit{delay increase}, \textit{number of cuts}, and \textit{interposer-routing interface}. We run all the experiments using the setup explained in Section~\ref{sec:CADeffect}.

%   - Impact of # of cutlines
\subsection{Impact of Number of Dice}

To examine the impact of the number of die used to construct an interposer-based FPGA, we varied the number of cuts from 1 to 3 (which varies the number of die from 2 to 4).  In these experiments, the \textit{delay increase} was kept constant at $1ns$.

@@

Figure~\ref{fig:standard_minW} shows the graph of minimum channel width versus \textit{\% wires cut}. It can be noted from this graph that the minimum channel width is almost steady up to 60\% of wires cut, indicating that the placement engine is able to avoid saturating the interposer routing until that point. When more than 60\% of the wires are cut however (i.e. the interposer provides less than 40\% of the usual within-die routing capacity), the minimum channel width grows rapidly indicating that the interposer routing bandwidth has become a limiting factor.

% TODO
% "extra FPGA routing channel width" vs. wires across the interposer
%\subsection{Intra-FPGA Routing  width vs. wires across the interposer}

%Can you take a crack at adding a section that tries to present the data as "extra FPGA routing channel width" vs. wires across the interposer?  The idea would be to address reviewer 4's comments:  the graph will show we need some minimum number of wires across the interposer, and it will also directly show how many more wires we need to add to the FPGA to maintain routability.  I think you could get this data simply by processing data you already have.
%So: 
%x-axis:  geomean wires crossing interposer
%y-axis:  geomean channel width required

Figure \ref{fig:crossingwires} provides an alternative way to visualize the relationship between interposer routing supply and routability. This figure shows how the geometric average minimum channel width required within the FPGA dice varies as the geometric average of the \emph{absolute} number of wires crossing the interposer in each channel varies, again for 2-die system. When 110 tracks cross the interposer, the interposer channels have the same capacity as the vertical routing channels within each FPGA die. As fewer wires cross the interposer, the channel width required within the FPGA dice increases to compensate for the routing difficulty in crossing the interposer. The increase is gradual as the interposer routing is reduced from 110 tracks per channel to 34 tracks per channel; over this range the routing per channel required in the FPGA dice increases from 110 tracks per channel to 114 tracks per channel. As the routing crossing the interposer is further reduced however, it becomes very difficult to increase the within-die routing sufficiently to compensate. At 18 tracks crossing the interposer channels, for example, the within-die routing must have a channel width of \emph{188} tracks to successfully route the designs. Clearly the CAD tools have the ability to trade-off interposer routing for within-die routing over a reasonable range but below a certain level (40\% of the original within-die minimum channel width in our experiments) routability becomes almost solely limited by the wiring crossing the interposer and further reduction in interposer routing is not productive.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{numberofcrossingwires_new.eps}
\caption{Geometric mean of required intra-die minimum channel width vs. geometric mean of the number of wires crossing the interposer for 2 dice and $1ns$ of \textit{delay increase}.}
\label{fig:crossingwires}
\end{figure}

To investigate the impact of the interposer delay (\textit{delay increase}), we keep all parameters constant and we sweep the \textit{delay increase} from 0 to 1.5 \textit{ns}. We use \textit{number of cuts} $= 1$ (corresponding to two dice), \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{delays_crit_path_new.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 2 dice and $0.0$, $0.5$, $1.0$ and $1.5ns$ of \textit{delay increase}.}
\label{fig:delays_crit}
\end{figure}

To measure the impact of routing-interposer interface parameters, we set \textit{delay increase}=$1ns$, \textit{number of cuts=1}, and \textit{\% wires cut=80\%}; then, we measure the impact of the \textit{bidirectional interposer wires}, \textit{fanin transfer}, and \textit{fanout transfer} parameters as well as different combinations of them. Figure~\ref{fig:all_arch_experiments} summarizes the results of these experiments.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{arch_experiments.eps}
\caption{Exploring more flexible interposer-based architectures by modifying routing-interposer interfaces.}
\label{fig:all_arch_experiments}
\end{figure}

Experiments show that using bidirectional interposer wires instead of unidirectional wires (column \textit{b} in Figure~\ref{fig:all_arch_experiments}) improves minimum channel width by 1\% and critical path delay by 0.5\%. Therefore, while it is beneficial to use bidirectional interposer wires, it does not dramatically improve performance. We believe that is because a well placed and routed design that fills the chip completely with little spare room has approximately equal number of signals crossing the interposer in each direction (i.e. number of signals transferred from die 1 to die 2 is close to the number of signals transferred from die 2 to die 1).

Adding multiplexers in front of interposer wires allows dangling routing wires to connect to interposer wires (\textit{``fanin transfer"}) and adding extra load on the output of the interposer wires (\textit{``fanout transfer"}) allows the interposer wire to drive dangling wires on the other side of the cut. Therefore these modifications provide higher flexibility to better utilize interposer wires. Experiments show that enabling these two options alone without using bidirectional interposer wires (column \textit{c} in Figure~\ref{fig:all_arch_experiments}) reduces minimum channel width by 3\% and reduces critical path delay by 1\%.

Enabling all of these architecture modifications simultaneously (column \textit{f} in Figure~\ref{fig:all_arch_experiments}) provides the highest level of flexibility for the router and achieves 11.2\% reduction in minimum channel width and 1.3\% reduction in critical path delay. These models show that interposer-based FPGAs can extensively benefit from hardware that provides a flexible routing-interposer interface.

The \textit{fanin transfer} and \textit{fanout transfer} options ensure that no vertical routing wires are dangling at the cutline, hence, each vertical routing wire either has exactly \textit{one} interposer wire driving it or it drives a mux of exactly \textit{one} interposer wire. As a final experiment, we try to add additional fanins and fanouts to interposer wires, so that each vertical routing wire can \textit{potentially} drive multiple interposer wires, or be driven by one of many interposer wires. We found that adding extra fanouts to interposer wires does not help performance or area, but adding additional fanins to interposer wire muxes helps reduce the minimum channel width while maintaining the same critical path delay. As shown in Figure~\ref{fig:additional_fanins}, adding 4 additional inputs to interposer wire muxes reduces the minimum channel width by 1\% and adding 8 additional inputs reduces minimum channel width by 1.5\%. After that point the performance is no longer limited by the ability to get on an interposer wire and hence we do not see any gains for adding more than 8 additional inputs. Notice that column \textit{0} of Figure~\ref{fig:additional_fanins} is the same as column \textit{f} of Figure~\ref{fig:all_arch_experiments}.

\section{Conclusions and Future Work}
\label{conclusionSection}
We have extended VPR to model and optimize for interposer-based multi-FPGA systems. While such systems are now a commercial reality, we do not know of any prior study of their key architectural parameters. We found that by modifying VPR's placement cost function we could improve the routability (reduce minW) by 18\%, while simultaneously improving speed by 2\%. Interestingly, we found that a smooth cost function that less precisely models the interposing wiring demand outperformed direct monitoring of the number of required interposer connections during annealing.

>>>>

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the ``flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}) \cite{luu2014vtr}. The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

@@

We believe the key to the good performance of the new placer cost functions is that \textit{cut\_cost} produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalize bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{bb_new.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

@@

\subsubsection{Effectiveness of router optimizations}
\label{sec:effectiveness_of_router_enhancements}
The router optimizations described in Section~\ref{sec:router_optimizations} greatly reduce the runtime of the router. Figure~\ref{fig:lookahead_runtime} shows the geomean of the router runtime for the 8 largest VTR circuits for different \textit{\% wires cut} values. On average, the router runs 31 times faster and finds identical minimum channel widths for all circuits.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt.eps}
\caption{Router runtime with original and updated lookahead function.}
\label{fig:lookahead_runtime}
\end{figure}

The new router lookahead changes the priority of the routes discovered by the router. Using the new router lookahead function increases the critical path delay by 1.3\% on average as shown in Figure~\ref{fig:lookahead_fmax}. We consider this small quality loss acceptable and use the enhanced router lookahead in all our experiments due to its greatly reduced runtime.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt_fmax.eps}
\caption{Impact of new vs. original router lookahead on the critical path delay.}
\label{fig:lookahead_fmax}
\end{figure}

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Adding Partitioning to the CAD Flow}
\label{sec:CADflow}
TODO: Javeed

@@

Using the best CAD settings found in Section~\ref{cadSection} and the best CAD flow described in Section~\ref{sec:CADflow}, we analyze the impact of the key architecture parameters: \textit{\% wires cut}, \textit{delay increase}, \textit{number of cuts}, and \textit{interposer-routing interface}. We run all the experiments using the setup detailed in Section~\ref{sec:CADeffect}.

To measure the impact of routing-interposer interface parameters, we set \textit{delay increase}=$1ns$, \textit{number of cuts=1}, and \textit{\% wires cut=80\%}; then, we measure the impact of the \textit{bidirectional interposer wires}, \textit{fanin transfer}, and \textit{fanout transfer} parameters in different combinations. Figure~\ref{fig:all_arch_experiments} summarizes the results of these experiments; all results are normalized to the least flexible routing-interposer interface.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{arch_experiments2.eps}
\caption{Impact of routing-interposer interface on minimum channel width and circuit speed.}
\label{fig:all_arch_experiments}
\end{figure}

Figure~\ref{fig:all_arch_experiments} shows that using bidirectional interposer wires instead of unidirectional wires (column \textit{b} in Figure~\ref{fig:all_arch_experiments}) improves minimum channel width by 1\% and critical path delay by 0.5\%. Therefore, while it is beneficial to use bidirectional interposer wires, in isolation they do not dramatically improve performance.

Adding multiplexers in front of interposer wires allows dangling routing wires to connect to interposer wires (\textit{``fanin transfer"}) and adding extra load on the output of the interposer wires (\textit{``fanout transfer"}) allows the interposer wire to drive dangling wires on the other side of the cut. Therefore these modifications provide higher flexibility to better utilize interposer wires. Figure~\ref{fig:all_arch_experiments} shows that enabling these two options alone without using bidirectional interposer wires (column \textit{c} in Figure~\ref{fig:all_arch_experiments}) reduces minimum channel width by 3\% and reduces critical path delay by 1\%.

Enabling all of these architecture modifications simultaneously (column \textit{f} in Figure~\ref{fig:all_arch_experiments}) provides the highest level of flexibility for the router and achieves 11.2\% reduction in minimum channel width and 1.3\% reduction in critical path delay. These models show that interposer-based FPGAs can extensively benefit from hardware that provides a flexible routing-interposer interface. It is important both to make the interposer wires easier to access with additional routing switches \textit{and} to make them bidirectional to accommodate asymmetric routing demand.

The \textit{fanin transfer} and \textit{fanout transfer} options ensure that no vertical routing wires are dangling at the cutline; hence, each vertical routing wire either has exactly \textit{one} interposer wire driving it or it drives a mux feeding exactly \textit{one} interposer wire. As a final experiment, we try to add additional fanins and fanouts to interposer wires, so that each vertical routing wire can \textit{potentially} drive multiple interposer wires, or be driven by one of many interposer wires. We found that adding these extra fanouts to interposer wires does not help performance or area, but adding additional fanins to interposer wire muxes helps reduce the minimum channel width while maintaining the same critical path delay. As shown in Figure~\ref{fig:additional_fanins}, adding 4 additional inputs to interposer wire muxes reduces the minimum channel width by 1\% and adding 8 additional inputs reduces minimum channel width by 1.5\%. After that point the performance is no longer limited by the ability to get on an interposer wire and hence we do not see any gains for adding more than 8 additional inputs. Notice that column \textit{0} of Figure~\ref{fig:additional_fanins} is the same as column \textit{f} of Figure~\ref{fig:all_arch_experiments}.

Clearly using bidirectional interposer wires with fanin transfer and fanout transfer is very beneficial, so we use these parameters in all future experiments. Adding additional fanin beyond transferring the fanin of cut wires is only a small further benefit so we use additional fanin=0 in subsequent sections. Adding these additional multiplexers will have a very small impact on die size as 10,000 muxes per cut would be typical in a large FPGA~\cite{xilinx7series} which is a very small fraction of the several million routing multiplexers in a large FPGA die.

Figure~\ref{fig:standard_minW} shows the graph of minimum channel width versus \textit{\% wires cut}. It can be noted from this graph that the minimum channel width is essentially constant up to 60\% of wires cut, indicating that the CAD flow is able to avoid saturating the interposer routing until that point. When more than 80\% of the wires are cut however (i.e. the interposer provides less than 20\% of the usual within-die routing capacity), the minimum channel width grows rapidly indicating that the interposer routing bandwidth has become a limiting factor.

Figure \ref{fig:crossingwires} provides an alternative way to visualize the relationship between interposer routing supply and routability. This figure shows how the geometric average minimum channel width required within the FPGA dice varies as the geometric average of the \emph{absolute} number of wires crossing the interposer in each channel varies, again for 2-die system. When 110 tracks cross the interposer, the interposer channels have the same capacity as the vertical routing channels within each FPGA die. As fewer wires cross the interposer, the channel width required within the FPGA dice increases to compensate for the routing difficulty in crossing the interposer. The increase is gradual as the interposer routing is reduced from 110 tracks per channel to 34 tracks per channel; over this range the routing per channel required in the FPGA dice increases from 110 tracks per channel to 114 tracks per channel. As the routing crossing the interposer is further reduced however, it becomes very difficult to increase the within-die routing sufficiently to compensate. At 18 tracks crossing the interposer channels, for example, the within-die routing must have a channel width of \emph{188} tracks to successfully route the designs. Clearly the CAD tools have the ability to trade-off interposer routing for within-die routing over a reasonable range but below a certain level (20\% of the original within-die minimum channel width in our experiments) routability becomes almost solely limited by the wiring crossing the interposer and further reduction in interposer routing is not productive.

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{numberofcrossingwires_new.eps}
\caption{Geometric mean of required intra-die minimum channel width vs. geometric mean of the number of wires crossing the interposer for 2 dice and $1ns$ of \textit{delay increase}.}
\label{fig:crossingwires}
\end{figure}

\begin{figure}[!b]
\centering
\includegraphics[width=\linewidth]{delays_crit_path_new.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 2 dice and $0.0$, $0.5$, $1.0$ and $1.5ns$ of \textit{delay increase}.}
\label{fig:delays_crit}
\end{figure}

To investigate the impact of the interposer delay (\textit{delay increase}), we keep all other interposer parameters constant and sweep the \textit{delay increase} from 0 to 1.5 \textit{ns}. We use \textit{number of cuts} $= 1$ (i.e. two dice), \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

%   - Impact of # of cutlines
\subsection{Impact of Number of Dice}
\label{num_dice_impact}
TODO: Javeed\\
To examine the impact of the number of die used to construct an interposer-based FPGA, we varied the number of cuts from 1 to 3 (which varies the number of die from 2 to 4).  In these experiments, the \textit{delay increase} was kept constant at $1ns$.

@@

%   - Impact of aspect ratio
\subsection{Impact of Aspect Ratio}
\label{aspect_ratio_impact}
TODO: Javeed

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Conclusions and Future Work}
\label{conclusionSection}
We have extended VPR to model and optimize for interposer-based multi-FPGA systems. While such systems are now a commercial reality, we do not know of any prior study of their key architectural parameters. We found that by modifying VPR's placement cost function we could improve the routability (reduce minW) by 18\%, while simultaneously improving speed by 2\%. Interestingly, we found that a smooth cost function that less precisely models the interposing wiring demand outperformed direct monitoring of the number of required interposer connections during annealing.
